---
title: "ERA5_20 year comparison"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(terra)
library(raster)
library(ncdf4)
library(lubridate)
library(sf)
library(tmap)

era5_vars<-read_csv("data/era5_vars.csv")
```

First, this is a python script for data. You have to have GEE set up as an account beforehand to authorize it. You can also use `pip install earthengine-api` on the Python command line to add the `ee` library.

```{python}
import ee
import datetime

# Authorize Earth engine (if needed)
# ee.Authenticate()

# Initialize Earth Engine
ee.Initialize()

# Define parameters
variables = [
    "dewpoint_temperature_2m",
    "temperature_2m_min",
    "temperature_2m_max",
    "temperature_2m",
    "total_precipitation_sum",
    "surface_pressure"
]

# Define date range
# start_date = datetime.date(1950, 1, 1)
# end_date = datetime.date(1950, 2, 1)

# Define region (global, you can restrict this to your area of interest)
region = ee.Geometry.Rectangle([-127, 23, -63, 50])

# Load ERA5-Land Daily Aggregated dataset
# dataset = ee.ImageCollection("ECMWF/ERA5_LAND/DAILY_AGGR") \
#    .filterDate(str(start_date), str(end_date))

# Export function
def export_image(image, variable_name, date_string):
    task = ee.batch.Export.image.toDrive(
        image=image.select(variable_name),
        description=f"{variable_name}_{date_string}",
        folder="GEE_ERA5_JanJul_cleanup",
        fileNamePrefix=f"{variable_name}_{date_string}",
        region=region,
        scale=10000,
        crs="EPSG:4326",
        maxPixels=1e13
    )
    task.start()
    print(f"Started export: {variable_name}_{date_string}")

# Loop over years and months (January and July)
for year in range(2021, 2024):
    for month in [1, 7]:  # January and July
        start_date = datetime.date(year, month, 1)
        if month == 1:
            end_date = datetime.date(year, 2, 1)
        else:  # July
            end_date = datetime.date(year, 8, 1)

        # Filter the collection for the month
        dataset = ee.ImageCollection("ECMWF/ERA5_LAND/DAILY_AGGR") \
            .filterDate(str(start_date), str(end_date))

        for var in variables:
            collection = dataset.select(var)
            size = collection.size().getInfo()

            for i in range(size):
                image = ee.Image(collection.toList(size).get(i))
                timestamp = image.date().format("YYYY-MM-dd").getInfo()
                export_image(image, var, timestamp)
```

#List downloaded files and download any that are missing The script above will export your files to a Google Drive folder. If that's different than this project folder, you'll need to move them over first.

Once you've done so, you can list the files and extract information on variable and date.

```{r}
files<-data.frame(files=list.files(path="data/raw_rasters",
                  full.names = T, recursive = T))%>%
  filter(!str_detect(files,".au")) %>%
  filter(!str_detect(files,".xml")) %>%
  separate(files,into=c("A","folder","file"),
           remove=F,sep="/") %>%
  dplyr::select(-A:-folder) %>%
  separate(file,into=c("varyear","monthday"),sep="-",
           extra="merge") %>%
  mutate(monthday=str_replace(monthday,".tif",""),
         year=substr(varyear,nchar(varyear)-3,nchar(varyear)),
         var=substr(varyear,1,nchar(varyear)-5)) %>%
  separate(monthday,into=c("month","day"),sep="-") %>%
  dplyr::select(-varyear) %>%
  mutate(fulldate=paste0(month,"-",day,"-",year),
         year_range=case_when(year<1955~"1950-1954",
                              year<1960~"1955-1959",
                              year<1965~"1960-1964",
                              year<1970~"1965-1969",
                              year>2002 & year<2009~"2004-2008",
                              year<2014~"2009-2013",
                              year<2019~"2014-2018",
                              year<2025~"2019-2024",
                              TRUE~"None"))
table(files$year_range)
```

Check for files that are missing. This is done in R.

```{r}
dates1<-data.frame(dates=seq(as.Date("1950/1/1"), as.Date("1950/1/31"), by = "day")) %>%
  mutate(md=paste0(substr(dates,6,7),"-",substr(dates,9,10))) %>%
  expand(md,year=1950:1969) %>%
  mutate(fulldate=paste0("M",md,"-",year)) %>%
  dplyr::select(-md,-year) %>%
  expand(fulldate,unique(files$var)) %>%
  rename(var=`unique(files$var)`) 

dates1a<-dates1 %>%
  bind_rows(dates1 %>%
              mutate(fulldate=str_replace(fulldate,"M01","M07")))
        

files_check<-dates1a %>%
  left_join(files %>%
              mutate(fulldate=paste0("M",fulldate))) %>%
  filter(is.na(files)) %>%
  distinct(fulldate)

#Later period
dates2<-data.frame(dates=seq(as.Date("1950/1/1"), as.Date("1950/1/31"), by = "day")) %>%
  mutate(md=paste0(substr(dates,6,7),"-",substr(dates,9,10))) %>%
  expand(md,year=2004:2024) %>%
  mutate(fulldate=paste0("M",md,"-",year)) %>%
  dplyr::select(-md,-year) %>%
  expand(fulldate,unique(files$var)) %>%
  rename(var=`unique(files$var)`)

dates2a<-dates2 %>%
  bind_rows(dates2 %>%
              mutate(fulldate=str_replace(fulldate,"M01","M07")))
        
files_check2<-dates2a %>%
  left_join(files %>%
              mutate(fulldate=paste0("M",fulldate))) %>%
  filter(is.na(files)) %>%
  distinct(fulldate)

files_check_all<-bind_rows(files_check,files_check2) %>%
  mutate(fulldate=str_replace(fulldate,"M",""))

```

Python script pt 2: cleanup and download missing files in Python

```{python}
import ee
import datetime
import pandas as pd

file_list = r.files_check_all

# Initialize Earth Engine
ee.Initialize()

# Define parameters
variables = [
    "dewpoint_temperature_2m",
    "temperature_2m_min",
    "temperature_2m_max",
    "temperature_2m",
    "total_precipitation_sum",
    "surface_pressure"
]

# Define region as the continental US
region = ee.Geometry.Rectangle([-127, 23, -63, 50])

# Export function
def export_image(image, variable_name, date_string):
    task = ee.batch.Export.image.toDrive(
        image=image.select(variable_name),
        description=f"{variable_name}_{date_string}",
        folder="GEE_ERA5_JanJul_cleanup",
        fileNamePrefix=f"{variable_name}_{date_string}",
        region=region,
        scale=10000,
        crs="EPSG:4326",
        maxPixels=1e13
    )
    task.start()
    print(f"Started export: {variable_name}_{date_string}")

# Example DataFrame with dates in 'files' column

# Convert the 'files' column to datetime objects
file_list['date'] = pd.to_datetime(file_list['fulldate'])

# Loop through each date in the DataFrame
for date in file_list['date']:
    start_date = date.strftime('%Y-%m-%d')
    end_date = (date + pd.Timedelta(days=1)).strftime('%Y-%m-%d')  # One day range

    # Filter the dataset for that specific day
    dataset = ee.ImageCollection("ECMWF/ERA5_LAND/DAILY_AGGR") \
        .filterDate(start_date, end_date)

    for var in variables:
        collection = dataset.select(var)
        size = collection.size().getInfo()

        for i in range(size):
            image = ee.Image(collection.toList(size).get(i))
            timestamp = image.date().format("YYYY-MM-dd").getInfo()
            export_image(image, var, timestamp)
            
```

#Import files

Now we can import files with terra and save them in 5 year batches.

```{r}
range_sel<-"1950-1954"

nc_create<-function(range_sel){
  
  batch_sel<-files %>%
    filter(year_range==range_sel)
  
  var_sel=vars[1]
  map(vars,function(var_sel){
    vars_char<-era5_vars %>%
      filter(var==var_sel)
    
    batch_sel_1<-batch_sel %>%
      filter(var==var_sel & month=="01")
      
    vardata_1<-rast(batch_sel_1$files)
    dates<-mdy((batch_sel_1$fulldate))

    # Add time to raster layers
    time(vardata_1) <- dates
    
    # Write to NetCDF with time metadata for January in range
    writeCDF(vardata_1,
           filename = paste0("data/",var_sel,"_jan_",unique(batch_sel_1$year_range),".nc"),
           varname = vars_char$var,
           unit = vars_char$var_unit,                     
           longname = vars_char$var_long,
           zname = "time",                 # dimension name
           compression=9,
           overwrite = TRUE)
  })
  
    map(vars,function(var_sel){
    vars_char<-era5_vars %>%
      filter(var==var_sel)
    
    batch_sel_1<-batch_sel %>%
      filter(var==var_sel & month=="07")
      
    vardata_1<-rast(batch_sel_1$files)
    dates<-mdy((batch_sel_1$fulldate))

    # Add time to raster layers
    time(vardata_1) <- dates
    
    # Write to NetCDF with time metadata for JULY
    writeCDF(vardata_1,
           filename = paste0("data/",var_sel,"_jul_",unique(batch_sel_1$year_range),".nc"),
           varname = vars_char$var,
           unit = vars_char$var_unit,                     
           longname = vars_char$var_long,
           zname = "time",                 # dimension name
           compression=9,
           overwrite = TRUE)
  })
}

map(unique(files$year_range),nc_create)
```

Manipulate the raster–calculate mean for the period and then do zonal statistics

```{r}
prcp1_b<-rast("dewpoint_temperature_2m_jul_1950-1954.nc") 

# Calculate mean value across years
prcp1_mean<-mean(prcp1_b)
plot(prcp1_mean)
prcp1_sd<-app(prcp1_b,fun=sd,na.rm=T)

states<-vect(tigris::states())
states <- project(states, crs(prcp1_mean))

zonal_means <- extract(prcp1_mean, states, fun = mean, na.rm = TRUE)
zonal_sd<-extract(prcp1_sd, states, fun = mean, na.rm = TRUE)

states$mean_temp<-zonal_means[,2]
states$mean_sd<-zonal_sd[,2]
states1<-st_as_sf(states) %>%
  filter(mean_temp>0)


tm_shape(states1)+tm_polygons("mean_temp")
tm_shape(states1)+tm_polygons("mean_sd")
```
